---
title: "final_Envs_193ds"
format: pdf
---

```{r}
#| message: false
#| warning: false
# loading packages
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(DHARMa) # checking diagnostics 
# reading in data
nest <- read_csv(here("data", "nest_data_final.csv"))
```

# Problem 1. Research writing

## a. Transparent statistical methods


**Part 1: generalized linear model**

Part 2: chi-square

- finding if there is a relationship between two categorical variables

## b. Suggestions for rewriting??

variable: native plant cover
## c. Further analysis

# Problem 2. Data analysis
## a. Clean and wrangle 
```{r}
# creating new object to clean and wrangle data
nests_clean <- nest |> 
  # filtering to only include ant species C. mimosae, C. sjostedti, and C. nigriceps
   filter(ant_species %in% c("RRB", "AB", "BBR")) |> 
  # creating new column with full scientific name
  mutate(species_name = recode_values(
    ant_species,
    "RRB" ~ "Crematogaster mimosae",
    "AB" ~ "Crematogaster sjostedti",
    "BBR" ~ "Crematogaster nigriceps"
    )) |> 
  # reordering ant species 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Crematogaster sjostedti",
                                    "Crematogaster mimosae",
                                    "Crematogaster nigriceps")) |>
# selecting only columns representing tree height, 
# if a bird nest is in tree, ant species, and ant species name
    select(height_cm, case_control, 
      species_name, ant_species) 
# checking structure of data frame
str(nests_clean)

slice_sample(nests_clean, # displaying nests clean data fram
             n = 10) # 10 rows
```


## b. Response variable

The 1s in the column case control means that the tree contained a nest and the 0s mean that there was an "available" for the birds to nest in. The researchers identified available trees by identifying the four nearest neighboring trees above 0.5 m tall to trees with nests. 

## c. Purpose of study

### Tree height

Whistling thorn tree height, measured in centimeters, is a continuous variable. The probability of bird nest occurrence would likely increase as tree height increases since there higher trees would be less accessible to ground-level predators. This information was found in paragraphs 3 and 4 of the introduction section. 


### Acacia Ant species

Acacia ant species is a categorical variable with three different leveles in our model (*C. mimosea*, *C. nigriceps*, and *C. sjostedti*). Each ant species provides different benefits to the tree and ecosystem. Ant species, like *C. mimosae* and *C. nigriceps*, that defend against herbivores greatly will probably increase the probability of nest occurrence as there are less large predators for the bird nests. This information was found in paragraphs 2 and 3 in the introduction section. 

**edit**
## d. Table of models

| Model Number | Tree Height | Ant Species |   Model Description              |
|--------------|:-----------:|:-----------:|:--------------------------------:|
|1             |      X      |      X      | all predictors (no interaction)  |
|2             |      X      |      X      | all predictors (interaction term)|           

## e. Run the models

```{r}
#| output: false
# model 1: no interaction, case control as a function of tree height and 
# ant species
model1 <- glm( 
  case_control ~ height_cm + species_name,
  data = nests_clean, # data frame: nests clean
  family = binomial) # binomial data

# model 2: interaction term, case control as a function of tree height and ant
# species
model2 <- glm(
  case_control ~ height_cm * species_name,
  data = nests_clean, # data frame: nests clean
  family = binomial) # binomial data
```


## f. Select the best model
```{r}
# using Akaike's Information Criterion (AIC) to choose best model
AICc(
  model1, # model 1
  model2 # model 2
) |> 
  arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) is the model with the interaction term between ant species and tree height as it explains the most variation in the nest occurrence in the simplest format.


```{r}
# displaying summary for best model
summary(model2)
```

## g. Check the diagnostics

```{r}
#| fig-width: 6
#| fig-height: 6
# checking diagnostics for model using DHARMa package
plot(simulateResiduals(model2))
```

From the residuals, the data is uniformally distributed as the plot follows the reference line and outliers do not have significant influence on the data since the lines follow the bands and are relatively flat. 

## h. Visualize the model predictions

```{r}
preds <- ggpredict(
  model2,
  terms = c("height_cm[all]", "species_name")) |> 
  rename(height_cm = x,
         species_name = group)
```

```{r}
# base layer: ggplot
ggplot(data = nests_clean,
       mapping = aes(x = height_cm,
                     y = case_control )) +
  geom_point(aes(color = species_name),
             shape = 21,
             alpha = 0.5) + 
  geom_line(data = preds,
            mapping = aes(x = height_cm,
                          y = predicted,
                          color = species_name),
            linewidth = 1) +
  geom_ribbon(data = preds,
              mapping = aes(
                        y = predicted,
                        ymin = conf.low,
                        ymax = conf.high,
                        fill = species_name),
               alpha = 0.1) +
  facet_wrap(~ species_name,
             labeller = labeller(
             species_name = c(
                 "Crematogaster sjostedti" = "C. sjostedti",
                 "Crematogaster mimosae"   = "C. mimosae",
                 "Crematogaster nigriceps" = "C. nigriceps"))) +
  scale_color_manual(values = c(
    "Crematogaster sjostedti" = "seagreen4",
    "Crematogaster mimosae" = "purple", 
    "Crematogaster nigriceps" = "skyblue")) +
  scale_fill_manual(values = c(
    "Crematogaster sjostedti" = "seagreen4",
    "Crematogaster mimosae"   = "purple",
    "Crematogaster nigriceps" = "skyblue")) +
scale_x_continuous(limits = c(0, 600), # set x axis to full range of tree height
                   breaks = c(0, 200, 400, 600)) +
labs (x = "Tree height (cm)",
      y = "Probability of nest occurrence") +
# changed theme from default
theme_classic() +
# removed legend
theme(legend.position = "none") +
# removing grid lines and grey background
theme(panel.grid = element_blank(),
      panel.background = element_blank())
```

## i. Write a caption for your figure.
*Figure 1.* **Probability of Nest Occurrence increases as tree height increases across all three ant species.** Points represent individual observations of nest occurrence (0 = no nest, 1 = nest present), with different colors representing different ant species. Lines represent predicted probability of nest occurrence using a binomial generalized linear model. Shaded ribbon represents 95% confidence interval around model predictions for each species. Data citation: Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna [Data set]. Zenodo. https://doi.org/10.5281/zenodo.8373322

## j. Calculate model predictions

```{r}
# calculating predicted probability of nest occurence at a value of 600 cm
ggpredict(model2,
          terms = c("height_cm [600]",
                    "species_name"))
```

## k. Interpret your results

Ant species and tree height interact to predict probability of nest occurrence. As Whistling thorn tree height increases, the probability of nest occurrence also increases. 

The effect of tree height on nest occurance probability is stronger for Crematogaster nigriceps 





# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

- For the affective visualization, I included all the different variables regarding my data versus the exploratory visualizations isolated two variables. Additionally, for the explanatory visualization, I displayed median and IQR but that was not displayed on my affective visualization. 

- Both of my visualizations utilize color to differentiate categorical variables such as work location. Additionally, both visualizations represent all 30 observations, just represented differently. 

- The main pattern that is visible in both visualizations is that when I worked at school, I had a higher productivity level than when I worked at any other place like the cafe or home. **FINISH THIS BASED ON AFFECTIVE VISUALIZATION**

- The main feedback I got in workshop was to make the variables more clear and be more clear with the overarching message of the data. I implemented this feedback by making the cloud size dependent on productivity level, so it was clear to see if there was a relationship between my variables and productivity levels. The decision to kept the suggestion was based on the fact that it helped convey the message behind the results better and was more clear for the reader. 

## b. Designing an analysis

-  correlation between sleep and productivity level
    
    

## c. Check any assumptions and run your analysis

## d. Create a visualization

## e. Write a caption

## f. Write about your results









